\chapter{Jobs}

A job is short-term and includes hourly pay. It may include some benefits. It is a way to earn money while you develop skills.

%\newcommand{\job}[6]{
%	\section*{#1}
%	\paragraph{Duties:} #6
%	\paragraph{Benefits:} #5
%	\paragraph{Payment Conditions:} \$#4 
%	\paragraph{Scheduling Conditions:} #3 
%	\paragraph{Requirements:} #2
%}

\job{Custodian}{Must be at least 4 years old.}{1 hour per day}{0.20 per hour per year old}
	{Gets priority on living spaces}{Cleans bathrooms and public spaces according to a schedule.}

\job{Gardener}{Must be 5 years old.}{3 hours per week}{0.22 per hour per year old}
	{Gets \$5 toy budget each month}{Helps with yard work and snow removal as needed.}

\job{Exterminator}{Must be 6 years old.}{As needed}{0.10 per pest killed responsibly}
	{Gets a \$1 base salary per week}{Kill harmful bugs, including roaches and spiders.}


\chapter{Careers}

A career is long-term and includes benefits and salaried pay.

\job{Landscaper}{Must be 10 years old with 6 months experience as a gardener.}{3 hours per week}{10 per month}
	{Bonus of \$200 toward transport (bike or car) after 3 months}{Completes yard work and snow removal as needed. Safely uses lawn mower and weed whacker in summer months.}

\job{I.T. Specialist}{Must be 12 years old with 1 year experience as a technician.}{On-call work}{18 per month}
	{Double screen time and a \$100 budget for new tech after 2 months}{Completes yard work and snow removal as needed. Safely uses lawn mower and weed whacker in summer months.}

\job{Babysitter}{Must be 12 years old with 1 year experience.}{1 night per week and as needed}{15 per month}{Choice of dinner on babysitting nights}{Babysit younger siblings during date nights and as needed}
